Depletion capabilities in the OpenMC Monte Carlo particle transport code

### 2020-05-21 manuscript.tex
Continued work on methodology

@@

\Cref{eq:depletion} simply states that the rate of change of $N_i$ is equal to
the production rate minus the loss rate. Because the equation for $N_i$ depends
on the density of other nuclides, $\{N_j \,|\, j \ne i\}$, \cref{eq:depletion}
represents a system of first-order differential equations. To form a proper
initial value problem, the nuclide densities at time 0 must be specified:
\begin{equation}
    N_i(0) = N_{i,0}.
\end{equation}
These equations can be written more compactly in matrix notation as
\begin{equation}
  \label{eq:depletion-matrix-t}
  \frac{d\mathbf{n}}{dt} = \mathbf{A}(\mathbf{n},t)\mathbf{n}, \quad \mathbf{n}(0) =
  \mathbf{n}_0
\end{equation}
where
\begin{equation}
  \mathbf{n} = \begin{pmatrix} N_1 \\ N_2 \\ \vdots \\ N_n \end{pmatrix}, \quad
  \mathbf{n}_0 = \begin{pmatrix} N_{1,0} \\ N_{2,0} \\ \vdots \\ N_{n,0} \end{pmatrix}
\end{equation}
and $\mathbf{A}(\mathbf{n},t) \in \mathbb{R}^{n\times n}$ is the burnup matrix
containing the decay and transmutation coefficients. Note that the burnup matrix
depends on $\mathbf{n}$ because the solution to the transport equation depends
on the nuclide densities, thereby making \cref{eq:depletion-matrix-t} nonlinear.

Although the neutron transport equation is time-dependent, the timescale over
which material compositions change is sufficiently long such that the transport
equation can be solved as a time-independent equation. Without loss of
generality, let us write the burnup matrix solely as a function of $\mathbf{n}$
so that \cref{eq:depletion-matrix-t} becomes
\begin{equation}
  \label{eq:depletion-matrix}
  \frac{d\mathbf{n}}{dt} = \mathbf{A}(\mathbf{n})\mathbf{n}, \quad \mathbf{n}(0) =
  \mathbf{n}_0.
\end{equation}
This better reflects the fact that the time, $t$, is generally not an input to
the transport equation, and hence the burnup matrix is uniquely determined by
the value of $\mathbf{n}$.

The principle difficulty of solving \cref{eq:depletion-matrix} is the fact that
the elements of $\mathbf{A}$ are not constant with respect to time. If
$\mathbf{A}$ is constant, \cref{eq:depletion-matrix} admits a closed-form
solution,
\begin{equation}
  \label{eq:constant-A}
  \mathbf{n}(t) = \exp \left (\mathbf{A} t \right ) \mathbf{n}_0.
\end{equation}
The exponential of a matrix $\mathbf{X}$ is defined by the power series
expansion
\begin{equation}
    \exp(\mathbf{X}) = \sum\limits_{k=0}^\infty \frac{1}{k!} \mathbf{X}^k
\end{equation}
where $\mathbf{X}^0 = \mathbf{I}$. Most numerical methods for solving
\cref{eq:depletion-matrix} involve two separate components:
\begin{enumerate}
  \item Approximating \cref{eq:depletion-matrix} in a form that involves taking
  one or more matrix exponentials, and
  \item Evaluating the matrix exponential or, equivalently, the action of a
  matrix exponential on a vector.
\end{enumerate}
We will henceforth refer to the first component as \emph{time integration}. Let
us now consider how each of these components are treated in OpenMC.

@@

Because an analytical solution to \cref{eq:depletion-matrix} is not available,
one must resort to using a numerical method that generates approximation
solutions in discrete increments. In general, the problem we need to solve then
is: given $\mathbf{n}_i \equiv \mathbf{n}(t_i)$, what is the value of
$\mathbf{n}_{i+1} \equiv \mathbf{n}(t_{i+1}) = \mathbf{n}(t_i + h)$, with $h$
being the step size? Without loss of generality, we shall choose $t_i=0$ as it
makes the presentation of various methods simpler.

@@

The simplest numerical method for solving \cref{eq:depletion-matrix} over a
timestep is known as the \emph{predictor} method. It proceeds by assuming that
$\mathbf{A}$ is constant over the timestep and assumes the value at the
beginning of the step. The solution then follows directly from
\cref{eq:constant-A}:
\begin{equation}
    \mathbf{n}_{i+1} = \exp\left(h\mathbf{A}(\mathbf{n}_i) \right) \mathbf{n}_i.
\end{equation}
The end-of-step composition is thus ``predicted'' by the reaction rates at the
beginning of the step. This method has the virtue of requiring a single
transport solution and a single matrix exponential but is only first-order
accurate.

@@

A series of \emph{predictor-corrector} methods that use multiple stages offer
improved accuracy over the predictor method. The simplest of these methods is
the CE/CM algorithm, which is the default algorithm used in
MCNP6~\citep{fensin2006tans}. In this method, $\mathbf{A}$ is evaluated at the
beginning of the step and used to deplete the composition over half the step.
Then, $\mathbf{A}$ is reevaluated using the predicted middle-of-step composition
and used to deplete over the full step. The algorithm can be expressed as
\begin{align}
    \mathbf{n}_{i+1/2} &= \exp \left (\frac{h}{2}\mathbf{A}(\mathbf{n}_i) \right) \mathbf{n}_i \\
    \mathbf{n}_{i+1} &= \exp \left(h \mathbf{A}(\mathbf{n}_{i+1/2}) \right) \mathbf{n}_i.
\end{align}
This method requires two transport solutions and two matrix exponential solves
(twice as expensive as the predictor method) and is second-order accurate.

More sophisticated predictor-corrector methods have been introduced by
\citet{isotalo2011ane2,isotalo2011ane3}. In the CE/LI method, the end-of-step
composition is first predicted using $\mathbf{A}(\mathbf{n}_i)$:
\begin{equation}
  \mathbf{n}_{i+1}^p = \exp \left( h\mathbf{A}(\mathbf{n}_i) \right) \mathbf{n}_i
\end{equation}
Then, $\mathbf{n}_{i+1}^p$, is used to reevaluate $\mathbf{A}$. An approximation
of $\mathbf{A}$ that depends on time is formed by linearly interpolating between
$\mathbf{A}(\mathbf{n}_i)$ and $\mathbf{A}\left(\mathbf{n}_{i+1}^p\right)$. This
contitutes a new differential equation for an approximation of $\mathbf{n}$ that
we shall denote $\hat{\mathbf{n}}$,
\begin{equation}
  \frac{d\hat{\mathbf{n}}}{dt} = \left [ \left ( 1 - \frac{t}{h} \right)
  \mathbf{A}(\mathbf{n}_i) + \frac{t}{h} \mathbf{A}\left(\mathbf{n}_{i+1}^p \right) \right ] \hat{\mathbf{n}}
\end{equation}

Our aim here is not to exhaustively describe all integration methods but rather
to give a few examples that elucidate the main considerations one must take into
account when choosing a method. Generally, there is a tradeoff between the
accuracy of the method and its computational expense. The expense is driven
almost entirely by the time to compute a transport solution, i.e., to evaluate
$\mathbf{A}$ for a given $\mathbf{n}$. Thus, the cost of a method scales with
the number of $\mathbf{A}$ evaluations that are performed per timestep. On the
other hand, methods that require more evaluations generally achieve higher
accuracy. The predictor method only requires one evaluation and its error
converges as $\mathcal{O}(h)$. The CE/CM method requires two evaluations and is
thus twice as expensive as the predictor method, but achieves an error of
$\mathcal{O}(h^2)$. An exhaustive description of time integration methods and
their merits can be found in the thesis of \citet{josey2017phd}.

@@

As we saw in \cref{sec:time_integration}, numerically integrating
\cref{eq:depletion-matrix} requires evaluating one or more matrix exponentials.
OpenMC uses the Chebyshev rational approximation method (CRAM), which was
introduced in a series of papers by Pusa~\citep{pusa2010nse,pusa2011nse}, to
evaluate matrix exponentials. In particular, OpenMC utilizes an incomplete
partial fraction (IPF) form~\citep{pusa2016nse} of CRAM that provides a good
balance of numerical stability and efficiency. In this representation the matrix
exponential is approximated as
\begin{equation}
    \exp(\mathbf{A}t) \approx \alpha_0 \prod\limits_{\ell=1}^{k/2} \left (
    \mathbf{I} + 2 \text{Re} \left ( \widetilde{\alpha}_\ell \left (\mathbf{A}t
    - \theta_\ell \mathbf{I} \right )^{-1} \right ) \right )
\end{equation}
where $k$ is the order of the approximation and $\alpha_0$,
$\widetilde{\alpha}_\ell$, and $\theta_\ell$ are coefficients that have been
tabulated for orders up to $k=48$. Rather than computing the full approximation
and then multiplying it by a vector, the following algorithm\footnote{The
original description of the algorithm presented by \citet{pusa2016nse} contains
a typo.} is used to incrementally apply the terms within the product:
\begin{algorithm}[H]
  \caption{Incomplete partial fraction form of CRAM.}
  \label{alg:cram}
  \begin{algorithmic}[1]
    \State $\mathbf{n} \gets \mathbf{n_0}$
    \For{$\ell \gets 1$ \textbf{to} $k/2$}
      \State $\mathbf{n} \gets \mathbf{n} + 2\text{Re}(\widetilde{\alpha}_\ell
        (\mathbf{A}t - \theta_\ell)^{-1})\mathbf{n}$
    \EndFor
    \State $\mathbf{n} \gets \alpha_0 \mathbf{n}$
  \end{algorithmic}
\end{algorithm}
The $k$th order approximation for CRAM requires solving $k/2$ sparse linear
systems. OpenMC relies on functionality from scipy.sparse.linalg for solving the
linear systems.

>>>>

\newcommand{\vect}[1]{\mathbf{#1}} % vectors and matrices

@@

\Cref{eq:depletion} simply states that the rate of change of $N_i$ is equal to
the production rate minus the loss rate. Because the equation for $N_i$ depends
on the density of other nuclides, $\{N_j \,|\, j \ne i\}$, \cref{eq:depletion}
represents a system of first-order ordinary differential equations (ODEs). To
form a proper initial value problem, the nuclide densities at time 0 must be
specified:
\begin{equation}
    N_i(0) = N_{i,0}.
\end{equation}
These equations can be written more compactly in matrix notation as
\begin{equation}
  \label{eq:depletion-matrix-t}
  \frac{d\vect{n}}{dt} = \vect{A}(\vect{n},t)\vect{n}, \quad \vect{n}(0) =
  \vect{n}_0
\end{equation}
where
\begin{equation}
  \vect{n} = \begin{pmatrix} N_1 \\ N_2 \\ \vdots \\ N_n \end{pmatrix}, \quad
  \vect{n}_0 = \begin{pmatrix} N_{1,0} \\ N_{2,0} \\ \vdots \\ N_{n,0} \end{pmatrix}
\end{equation}
and $\vect{A}(\vect{n},t) \in \mathbb{R}^{n\times n}$ is the burnup matrix
containing the decay and transmutation coefficients. Note that the burnup matrix
depends on $\vect{n}$ because the solution to the transport equation depends
on the nuclide densities, thereby making \cref{eq:depletion-matrix-t} nonlinear.

Although the neutron transport equation is time-dependent, the timescale over
which material compositions change is sufficiently long such that the transport
equation can be solved as a time-independent equation. Without loss of
generality, let us write the burnup matrix solely as a function of $\vect{n}$
so that \cref{eq:depletion-matrix-t} becomes
\begin{equation}
  \label{eq:depletion-matrix}
  \frac{d\vect{n}}{dt} = \vect{A}(\vect{n})\vect{n}, \quad \vect{n}(0) =
  \vect{n}_0.
\end{equation}
This better reflects the fact that the time, $t$, is generally not an input to
the transport equation, and hence the burnup matrix is uniquely determined by
the value of $\vect{n}$.

The principle difficulty of solving \cref{eq:depletion-matrix} is the fact that
the elements of $\vect{A}$ are not constant with respect to time. If
$\vect{A}$ is constant, \cref{eq:depletion-matrix} admits a closed-form
solution,
\begin{equation}
  \label{eq:constant-A}
  \vect{n}(t) = \exp \left (\vect{A} t \right ) \vect{n}_0.
\end{equation}
The exponential of a matrix $\vect{X}$ is defined by the power series
expansion
\begin{equation}
    \exp(\vect{X}) = \sum\limits_{k=0}^\infty \frac{\vect{X}^k}{k!}
\end{equation}
where $\vect{X}^0 = \vect{I}$. Most numerical methods for solving
\cref{eq:depletion-matrix} involve two separate components:
\begin{enumerate}
  \item Approximating \cref{eq:depletion-matrix} in a form that involves taking
  one or more matrix exponentials, and
  \item Evaluating the matrix exponential or, equivalently, the action of a
  matrix exponential on a vector.
\end{enumerate}
We will henceforth refer to the first component as \emph{time integration}. Let
us now consider how each of these components are treated in OpenMC.

@@

Because an analytical solution to \cref{eq:depletion-matrix} is not available,
one must resort to using a numerical method that generates approximation
solutions in discrete increments. In general, the problem we need to solve then
is: given $\vect{n}_i \equiv \vect{n}(t_i)$, what is the value of
$\vect{n}_{i+1} \equiv \vect{n}(t_{i+1}) = \vect{n}(t_i + h)$, with $h$
being the step size? Without loss of generality, we shall choose $t_i=0$ as it
makes the presentation of various methods simpler.

@@

The simplest numerical method for solving \cref{eq:depletion-matrix} over a
timestep is known as the \emph{predictor} method. It proceeds by assuming that
$\vect{A}$ is constant over the timestep and assumes the value at the
beginning of the step. The solution then follows directly from
\cref{eq:constant-A}:
\begin{equation}
    \vect{n}_{i+1} = \exp\left(h\vect{A}(\vect{n}_i) \right) \vect{n}_i.
\end{equation}
The end-of-step composition is thus ``predicted'' by the reaction rates at the
beginning of the step. This method has the virtue of requiring a single
transport solution and a single matrix exponential per timestep but is only
first-order accurate.

@@

A series of \emph{predictor-corrector} methods that use multiple stages offer
improved accuracy over the predictor method. The simplest of these methods is
the CE/CM algorithm, which is the default algorithm used in
MCNP6~\citep{fensin2006tans}. In this method, $\vect{A}$ is evaluated at the
beginning of the step and used to deplete the composition over half the step.
Then, $\vect{A}$ is reevaluated using the predicted middle-of-step composition
and used to deplete over the full step. The algorithm can be expressed as
\begin{equation}
  \begin{split}
    \vect{n}_{i+1/2} &= \exp \left (\frac{h}{2}\vect{A}(\vect{n}_i) \right) \vect{n}_i \\
    \vect{n}_{i+1} &= \exp \left(h \vect{A}(\vect{n}_{i+1/2}) \right) \vect{n}_i.
  \end{split}
\end{equation}
This method requires two transport solutions and two matrix exponential solves
per timestep (twice as expensive as the predictor method) and is second-order
accurate.

More sophisticated predictor-corrector methods have been introduced by
\citet{isotalo2011ane2,isotalo2011ane3}. In the CE/LI method, the end-of-step
composition is first predicted using $\vect{A}(\vect{n}_i)$:
\begin{equation}
  \vect{n}_{i+1}^p = \exp \left( h\vect{A}(\vect{n}_i) \right) \vect{n}_i.
\end{equation}
Then, $\vect{n}_{i+1}^p$ is used to reevaluate $\vect{A}$. An approximation of
$\vect{A}$ that depends on time is formed by linearly interpolating between
$\vect{A}(\vect{n}_i)$ and $\vect{A}\left(\vect{n}_{i+1}^p\right)$.
\Cref{eq:depletion-matrix} thus becomes
\begin{equation}
  \label{eq:linear-ode}
  \frac{d\vect{n}}{dt} = \hat{\vect{A}}(t) \vect{n}
\end{equation}
where
\begin{equation}
  \label{eq:linear-interp}
  \hat{\vect{A}}(t) = \left ( 1 - \frac{t}{h} \right) \vect{A}(\vect{n}_i) +
  \frac{t}{h} \vect{A}\left(\vect{n}_{i+1}^p \right).
\end{equation}
We see that $\hat{\vect{A}}$ no longer depends on $\vect{n}$ thereby making
\cref{eq:linear-ode} a linear nonautonomous ODE. Because $\hat{\vect{A}}$ is a
function of time, an approximation is still needed to obtain $\vect{n}_{i+1}$.
The simplest approach~\citep{isotalo2011ane2} is to average
\cref{eq:linear-interp} over the step,
\begin{equation}
  \bar{\vect{A}} = \int_0^h dt \, \hat{\vect{A}}(t) = \frac{\vect{A}(\vect{n}_i)
  + \vect{A}(\vect{n}_{i+1}^p)}{2},
\end{equation}
and then use $\bar{\vect{A}}$ to deplete over the full step:
\begin{equation}
  \vect{n}_{i+1} = \exp \left ( h \bar{\vect{A}} \right ) \vect{n}_i.
\end{equation}
\citet{isotalo2011ane3} suggested an improvement over this approach wherein
$\hat{\vect{A}}$ is averaged over substeps instead of the entire interval. That
is, if we divide the timestep into $m$ substeps, the solution is given by
\begin{equation}
  \begin{split}
    \label{eq:celi-substeps}
    \vect{A}_s &= \int_{(s-1)h/m}^{sh/m} dt \; \hat{\vect{A}}(t), \quad s=1,\dots,m \\
    \vect{n}_{i+1} &= \exp\left(\vect{A}_m \right) \exp \left(\vect{A}_{m-1}\right)
    \cdots \exp\left(\vect{A}_1\right) \vect{n}_i.
  \end{split}
\end{equation}
While using substeps is a reasonable approach for solving
\cref{eq:linear-ode}, there are other methods as well. For example, some
codes use a backward differentiation formula rather than a method based on an
exponential integrator~\cite{carpenter2009mc,hykes2017mc,sublet2017nds}.

\subsubsection{Magnus expansion integrators}

Any linear approximation of \cref{eq:depletion-matrix} characterized by
$\vect{A}$ that depends only on time, such as \cref{eq:linear-ode}, has an
infinite series solution known as the Magnus expansion~\citep{blanes2009pr}.
This expansion takes the form
\begin{equation}
  \begin{split}
  \vect{n}(t) &= \exp \left( \vect{\Omega}(t) \right ) \vect{n}_0 \\
  \vect{\Omega}(t) &= \sum\limits_{k=1}^\infty \vect{\Omega}_k(t).
  \end{split}
\end{equation}
The first three terms of the expansion are given by
\begin{equation}
  \label{eq:magnus-terms}
  \begin{split}
    \vect{\Omega}_1(t) &= \int_0^t dt_1 \vect{A}(t_1) \\
    \vect{\Omega}_2(t) &= \int_0^t dt_1 \int_0^{t_1} dt_2 [\vect{A}(t_1), \vect{A}(t_2) ] \\
    \vect{\Omega}_3(t) &= \int_0^t dt_1 \int_0^{t_1} dt_2 \int_0^{t_2} dt_3
      \left ( [\vect{A}(t_1), [\vect{A}(t_2), \vect{A}(t_3)]] + [\vect{A}(t_3), [\vect{A}(t_2), \vect{A}(t_1)]] \right )
  \end{split}
\end{equation}
where $[\vect{A},\vect{B}] = \vect{A}\vect{B} - \vect{B}\vect{A}$ is a matrix
commutator. Subsequent terms can be defined through a recurrence relation.

The Magnus expansion can be used as the basis for developing numerical
integration algorithms and has been widely studied. The typical approach is to
truncate the $\vect{\Omega}$ series at an appropriate order and apply an
approximation to the multivariate integrals in \cref{eq:magnus-terms}. In
particular, OpenMC uses a fourth-order commutator-free exponential
integrator~\citep{blanes2006anm} that is derived from the Magnus expansion in
order to solve \cref{eq:linear-ode}. In this approximation, the Magnus expansion
is truncated at two terms and then expressed as the composition of two
exponentials. For a system of ODEs in the form of \cref{eq:linear-ode}, the
resulting method is
\begin{equation}
  \label{eq:cf4}
  \vect{n}_{i+1} = \exp (a_2 \vect{A}_1 + a_1 \vect{A}_2)
  \exp (a_1 \vect{A}_1 + a_2 \vect{A}_2) \vect{n}_i
\end{equation}
where $\vect{A}_1$ and $\vect{A}_2$ correspond $\hat{\vect{A}}$ evaluated at the
nodes of a two-point Gauss-Legendre quadrature over the interval $[0,h]$:
\begin{equation}
  \begin{split}
    a_1 &= \frac{1}{4} + \frac{\sqrt{3}}{6}, \; a_2 = \frac{1}{4} - \frac{\sqrt{3}}{6} \\
    c_1 &= \frac{1}{2} - \frac{\sqrt{3}}{6}, \; c_2 = \frac{1}{2} + \frac{\sqrt{3}}{6} \\
    \vect{A}_1 &= h\hat{\vect{A}}\left ( c_1 h\right ), \; \vect{A}_2 = h\hat{\vect{A}}\left ( c_2 h\right ).
  \end{split}
\end{equation}
Evaluating $\vect{A}_1$ and $\vect{A}_2$ using \cref{eq:linear-interp} and
substituting in \cref{eq:cf4}, we obtain a CE/LI method based on the
fourth-order commutator-free Magnus integrator:
\begin{equation}
  \vect{n}_{i+1} = \exp \left( \frac{h}{12} \vect{A}(\vect{n}_i) + \frac{5h}{12} \vect{A}(\vect{n}_{i+1}^p) \right)
  \exp \left( \frac{5h}{12} \vect{A}(\vect{n}_i) + \frac{h}{12} \vect{A}(\vect{n}_{i+1}^p) \right) \vect{n}_i.
\end{equation}

OpenMC does include
both CE/LI and LE/QI methods~\cite{isotalo2011ane2}, but the approach for
solving \cref{eq:linear-ode} is a bit different.

Our aim here is not to exhaustively describe all integration methods but rather
to give a few examples that elucidate the main considerations one must take into
account when choosing a method. Generally, there is a tradeoff between the
accuracy of the method and its computational expense. The expense is driven
almost entirely by the time to compute a transport solution, i.e., to evaluate
$\vect{A}$ for a given $\vect{n}$. Thus, the cost of a method scales with
the number of $\vect{A}$ evaluations that are performed per timestep. On the
other hand, methods that require more evaluations generally achieve higher
accuracy. The predictor method only requires one evaluation and its error
converges as $\mathcal{O}(h)$. The CE/CM method requires two evaluations and is
thus twice as expensive as the predictor method, but achieves an error of
$\mathcal{O}(h^2)$. An exhaustive description of time integration methods and
their merits can be found in the thesis of \citet{josey2017phd}.

@@

As we saw in \cref{sec:time_integration}, numerically integrating
\cref{eq:depletion-matrix} requires evaluating one or more matrix exponentials.
OpenMC uses the Chebyshev rational approximation method (CRAM), which was
introduced in a series of papers by Pusa~\citep{pusa2010nse,pusa2011nse}, to
evaluate matrix exponentials. In particular, OpenMC utilizes an incomplete
partial fraction (IPF) form~\citep{pusa2016nse} of CRAM that provides a good
balance of numerical stability and efficiency. In this representation the matrix
exponential is approximated as
\begin{equation}
    \exp(\vect{A}t) \approx \alpha_0 \prod\limits_{\ell=1}^{k/2} \left (
    \vect{I} + 2 \text{Re} \left ( \widetilde{\alpha}_\ell \left (\vect{A}t
    - \theta_\ell \vect{I} \right )^{-1} \right ) \right )
\end{equation}
where $k$ is the order of the approximation and $\alpha_0$,
$\widetilde{\alpha}_\ell$, and $\theta_\ell$ are coefficients that have been
tabulated for orders up to $k=48$. Rather than computing the full approximation
and then multiplying it by a vector, the following algorithm\footnote{The
original description of the algorithm presented by \citet{pusa2016nse} contains
a typo.} is used to incrementally apply the terms within the product:
\begin{algorithm}[H]
  \caption{Incomplete partial fraction form of CRAM.}
  \label{alg:cram}
  \begin{algorithmic}[1]
    \State $\vect{n} \gets \vect{n_0}$
    \For{$\ell \gets 1$ \textbf{to} $k/2$}
      \State $\vect{n} \gets \vect{n} + 2\text{Re}(\widetilde{\alpha}_\ell
        (\vect{A}t - \theta_\ell)^{-1})\vect{n}$
    \EndFor
    \State $\vect{n} \gets \alpha_0 \vect{n}$
  \end{algorithmic}
\end{algorithm}
The $k$th order approximation for CRAM requires solving $k/2$ sparse linear
systems. OpenMC relies on functionality from scipy.sparse.linalg for solving the
linear systems.
